BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 implementation.tex
Second Batch of fixes (Most of Daniele's)

\subsection{Intel Pin}
BlueTracer is implemented using Pin, a dynamic binary instrumentation framework by Intel, which is vastly utilized for program analysis, testing of software and in the security field. The version of Pin used for the development of the tool is 3.5, in order to benefit from the notable improvements, both in terms of execution speed and offered features, which were introduced going from the 2.14 release to the 3.x series. Pin comes with its own OS-agnostic and compiler-agnostic runtime, called PinCRT. PinCRT exposes three layers: a generic operating system interface providing basic OS services (e.g. process and thread control), together with C and C++03 (without RTTI) runtime layers \cite{Pin}.

@@

\section{Thread Management}

@@

\subsection{Log Files and Multithreading}

@@

\section{Native APIs Tracing}
\iffalse
Talk about log structure when talking about how arguments are printed
\fi

Windows Native APIs are employed to call operating system services in kernel mode in a controlled way.  
In fact, all core Windows components, which possess direct access to hardware and services in charge of handling the computer's resources (e.g. memory), operate in kernel mode. This means that, every time a user mode application desires to carry out certain actions, like for instance starting a thread or allocating virtual memory, they must rely on kernel mode  services. The Windows Native API corresponds to the system call interface of standard monolithic operating systems, such as the majority of UNIX-like systems, with the difference that in the latter case the system call interface is documented and can be utilized directly by applications. Instead, due to Windows' architecture, Windows Native APIs are concealed to the programmer by the higher level Windows (\texttt{Win32}) APIs \cite{NTAPI1}. User mode Windows Native APIs, which are identified by their \texttt{Nt} prefix and are exported by \texttt{ntdll}, have caught the attention of malware writers since they are seen as a way of bypassing the documented APIs with the objective of performing a series of actions without being discovered \cite{NTAPI2}. For this reason it is a good idea to also trace them, in addition to the ordinary Windows APIs.

@@

\subsection{Main Image Check}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether or not only Native APIs called invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure~\ref{NTAPI Tracing}}, the first thing which is done in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

Having obtained the main executable address range in this way, it is then employed in \texttt{TraceSysCallEntry} to learn if the Native API call is occuring directly from the main image. Internally, before entering kernel mode every Native API executes some common code, i.e. each Native API stores its ordinal in the \texttt{eax} register and invokes \texttt{KiFastSystemCall}, where the \texttt{sysenter} instruction is used to actually enter kernel mode. When \texttt{sysenter} is executed, the kernel obtains the ordinal number from \texttt{eax} and utilizes it to call the corresponding function, prior to going back to user mode \cite{MalwareTech}.

@@

\begin{enumerate}
\item The first check involves discovering whether or not \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item If the first check fails, the second one is carried out, in which it is examined if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure~\ref{sysenter}}, such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

\subsection{Native API Name Resolution}

@@

\subsection{Native API Information Representation and Retrieval}

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether or not all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{struct}s (\textit{Listing 3.3}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e. 18), since a Native API can have at most 18 arguments. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure~\ref{NTAPI Tracing}}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building an hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Native API Logging}
With reference to \textit{Figure~\ref{NTAPI Tracing}}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed, i.e., in \texttt{TraceSysCallEntry}, and set to false when the logging is occurring after the Native API is executed, i.e. in \texttt{TraceSysCallExit}.
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}
\begin{lstlisting}[caption={\texttt{struct} representing an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {

@@

As previously mentioned in section 3.1.1, each thread possesses its own log file, whose pointer resides in the thread local storage, and writes on it by means of the \texttt{file\_write} function, also located in the thread local storage (\textit{Listing 3.1}). The first Native API component being logged is its name and this is done with the following format:
\\ 
\texttt{\textasciitilde\textasciitilde[System ID of thread]\textasciitilde\textasciitilde} \texttt{ [Call Counter] [Image Name]![Native API Name]}  
In the context of Native APIs, the name of the image is always \texttt{ntdll}, but this is not the case when tracing APIs.

@@

\begin{itemize}
\item Firstly, it is determined if the argument being logged is a named constant, i.e., it is checked if \texttt{enum\_name} is different from \texttt{NULL}. In that case, the \texttt{get\_arg\_syname} function is invoked, which is in charge of, given the input named constant value, finding the corresponding name and recording it to the log. The named constant resolution is carried out by initially fetching the \texttt{struct} containing all the members of the enumeration the named constant belongs to, through the use of the aforementioned hash map. Then, it is determined if there is a perfect match between the argument value and one of the values of the \texttt{struct}'s entries. If this occurs, then the associated symbolic name is simply retrieved from the \texttt{struct}'s entry. Otherwise, a linear search is performed to unveil possible composite named constants (e.g. \texttt{FILE\_SHARED\_READ | FILE\_SHARED\_WRITE}).

@@

\begin{itemize}
\item Then, if the argument's type is primitive (e.g. \texttt{int}, \texttt{bool}, etc.),  \texttt{print\_simple\_value} is invoked to record its value. In particular, the second parameter of this function is a boolean which is set to \texttt{true} if it is wanted to print the argument's value with leading zeros (e.g. if the argument is of pointer type and the address it stores is being logged ) and \texttt{false} otherwise. The way \texttt{print\_simple\_value} operates is quite straightforward:
\begin{enumerate}
\item At the beginning the argument's value is simply logged, with leading zeroes or not depending on the previously mentioned parameter
\item Then, if the argument is a pointer, it is determined if the value it points to needs to be logged. This occurs if the argument is being processed before the Native API execution (i.e. \texttt{pre} is \texttt{true}) and is an input parameter (i.e., \texttt{SYSARG\_READ} is contained in \texttt{flags}) or if the argument is being processed after the Native API execution (i.e., \texttt{pre} is \texttt{false}) and is an output parameter (i.e., \texttt{SYSARG\_WRITE} is contained in \texttt{flags}). The pointed value is obtained through the use of \texttt{PIN\_SafeCopy}, which copies the specified number of bytes from a source memory region to a destination memory region, guaranteeing safe return to the caller even if such regions are inaccessible.  
\end{enumerate}

@@

Once the input argument values are logged, then the reference to the \texttt{syscall\_tracer struct} representing the Native API under analysis is stored in thread local data (\textit{Listing 3.1}). This is done so that, after the Native API's execution, the corresponding \texttt{syscall\_tracer} data structure can be accessed from \texttt{TraceSysCallExit}. In fact, as mentioned in section 3.2, the \texttt{bluepill\_tls struct} associated to the running thread can be retrieved in any analysis function by means of \texttt{PIN\_GetThreadData}.

@@

\section{APIs Tracing}

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamic-link libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g. file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a very high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\label{API Tracing}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure~\ref{API Tracing}}.

@@

\subsection{APIs Instrumentation}

The first step taken towards APIs tracing is APIs instrumentation, i.e., placing analysis routines before and after the APIs' execution. To do this, a specific function was embedded in BluePill's image callback (the notion of image callback was previously introduced in section 3.3.1). Such function, named \texttt{AddAnalysisRoutines}, performs the following actions for every DLL being loaded:

\begin{enumerate} 
\item It inspects the symbol table, i.e., the DLL's exports. To this end, Pin symbol objects (\texttt{SYM}s) are employed, which provide information about function symbols in the applications \cite{Pin}.
\item For every symbol object, it retrieves the corresponding routine (\texttt{RTN}) object. In fact, in Pin a \texttt{RTN} represents a function usually generated by a compiler for a procedural programming language, e.g C \cite{Pin}. To retrieve the aforementioned \texttt{RTN} object, the \texttt{RTN\_FindByAddress} API was utilized, which takes as input a \texttt{RTN} memory address and returns an handle to the found \texttt{RTN}. Such memory address was determined by adding the symbol's RVA (obtained via \texttt{SYM\_Value}) to the lowest image address (fetched through \texttt{IMG\_LowAddress}).  
\item For every routine object, it retrieves the corresponding name via \texttt{RTN\_name} and checks whether or not the API being represented is "known", i.e. a source of information associated to the API exists.

@@

\textit{drltrace} (already mentioned in section 2.3) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry on \textit{Listing 3.9}.      
\vspace{1cm}
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

Before API instrumentation takes place, BlueTracer parses the aforementioned configuration file and, for every entry, builds a \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) for each argument in the entry. It was decided to reuse \texttt{sysinfo\_arg\_t struct} since its fields were fitting with respect to the arguments information which had to be stored. Therefore, after this parsing process, each configuration file entry is represented by an <API name, \texttt{vector} of \texttt{sysinfo\_arg\_t}> pair. Such pairs are then utilized to build an hash map, which maps the API name to the corresponding arguments. This hash map is used in step 3 of the APIs instrumentation process to, given the API name, fetch the appropriate information. In particular, it is important to note that, from a performance point of view, it is much better to include the information retrieval operations in the image instrumentation function, rather than in the analysis ones, as the former are executed just once for every image, while the latter are invoked each time an API is called by the pinned application.

The other source of API information employed by BlueTracer is \textit{PyREbox}'s database. \textit{PyREbox} is a Python scriptable reverse engineering sandbox based on QEMU and VMI techniques, the goal of which is to provide support for reverse engineering through its dynamic analysis capabilities \cite{Pyrebox}. \textit{PyREbox} is equipped with its own API tracer, which relies on the information contained in a sqlite database to correctly log the APIs' parameters. Such database was, in turn, generated by utilizing as its core the data from the \textit{Deviare}\footnote{Deviare is an open-source hooking engine for instrumenting arbitrary \texttt{Win32} functions \cite{Deviare}} project and then employing a crawler for parsing the Microsoft online documentation with the goal of marking which parameters are input parameters and which are output parameters. The resulting database is incredibly richer in API information than \textit{drltrace}'s configuration file. In fact, the number of APIs for which data is stored is much higher, argument names are also included, as well as the contents of named constants' enumerations, information which \textit{drltrace}'s configuration file does not provide at all.

@@

The most important field of such \texttt{struct} is surely the arguments' array, which is made up of \texttt{libcall\_arg\_info\_t} data structures (\textit{Listing 3.11}) containing the information relative each argument of an API, including an entry specific for the return value data. The array size was set to 19, since it was observed from the database data that the API with the most parameters had 18 arguments. This time, it was decided not to reuse NTAPI's \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) to store the API's arguments information due to its lack of the field storing the argument name. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing information associated to an API argument},captionpos=b]
typedef struct _libcall_arg_info_t {
	int arg_num;				// Arg Ordinal
	const char* arg_name;		// Arg Name
	int arg_type;             	// Arg Type
	char* arg_type_name;		// Arg Type Name
	int size;					// Size
	bool in_out_flag;			// Input/Output Flag
} libcall_arg_info_t;
\end{lstlisting}

@@

Due to the high amount of APIs for which information was stored in the \textit{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure~\ref{Hashmap}}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\label{Hashmap}
\end{figure}

@@

\subsection{API Analysis before Execution}

>>>>

\subsection{Intel Pin}
BlueTracer is implemented using Pin, a dynamic binary instrumentation framework by Intel, which is vastly utilized for program analysis, testing of software and in the security field. The version of Pin used for the development of the tool is 3.5, in order to benefit from the notable improvements, both in terms of execution speed and offered features, which were introduced going from the 2.14 release to the 3.x series. Pin comes with its own OS-agnostic and compiler-agnostic runtime, called PinCRT. PinCRT exposes three layers: a generic operating system interface providing basic OS services (e.g., process and thread control), together with C and C++03 (without RTTI) runtime layers \cite{Pin}.

@@

\section{Thread Management} \label{ThreadMan}

@@

\subsection{Log Files and Multithreading} \label{LogFilesMultithreading}

@@

\section{Tracing System Calls}
\iffalse
Talk about log structure when talking about how arguments are printed
\fi

Windows Native APIs are employed to call operating system services in kernel mode in a controlled way.  
In fact, all core Windows components, which possess direct access to hardware and services in charge of handling the computer's resources (e.g., memory), operate in kernel mode. This means that, every time a user mode application desires to carry out certain actions, like for instance starting a thread or allocating virtual memory, they must rely on kernel mode  services. The Windows Native API corresponds to the system call interface of standard monolithic operating systems, such as the majority of UNIX-like systems, with the difference that in the latter case the system call interface is documented and can be utilized directly by applications. Instead, due to Windows' architecture, Windows Native APIs are concealed to the programmer by the higher level Windows (\texttt{Win32}) APIs \cite{NTAPI1}. User mode Windows Native APIs, which are identified by their \texttt{Nt} prefix and are exported by \texttt{ntdll}, have caught the attention of malware writers since they are seen as a way of bypassing the documented APIs with the objective of performing a series of actions without being discovered \cite{NTAPI2}. For this reason it is a good idea to also trace them, in addition to the ordinary Windows APIs.

@@

\subsection{Main Image Check} \label{MainImage}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether only Native APIs invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure~\ref{NTAPI Tracing}}, the first thing we do in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

Having obtained the main executable address range in this way, it is then employed in \texttt{TraceSysCallEntry} to learn if the Native API call is occuring directly from the main image. Internally, before entering kernel mode every Native API executes some common code, i.e., each Native API stores its ordinal in the \texttt{eax} register and invokes \texttt{KiFastSystemCall}, where the \texttt{sysenter} instruction is used to actually enter kernel mode. When \texttt{sysenter} is executed, the kernel obtains the ordinal number from \texttt{eax} and utilizes it to call the corresponding function, prior to going back to user mode \cite{MalwareTech}.

@@

\begin{enumerate}
\item The first check involves discovering whether \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item If the first check fails, the second one is carried out, in which it is examined if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure~\ref{sysenter}}, such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

\subsection{Name Resolution}

@@

\subsection{Information Representation and Retrieval}

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{struct}s (\textit{Listing 3.3}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e., 18), since a Native API can have at most 18 arguments. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure~\ref{NTAPI Tracing}}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building a hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Logging}
With reference to \textit{Figure~\ref{NTAPI Tracing}}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed, i.e., in \texttt{TraceSysCallEntry}, and set to false when the logging is occurring after the Native API is executed, i.e., in \texttt{TraceSysCallExit}.
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}
\begin{lstlisting}[caption={\texttt{struct} representing an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {

@@

As previously mentioned in Section~\ref{LogFilesMultithreading}, each thread possesses its own log file, whose pointer resides in the thread local storage, and writes on it by means of the \texttt{file\_write} function, also located in the thread local storage (\textit{Listing 3.1}). The first Native API component being logged is its name and this is done with the following format:
\\ 
\texttt{\textasciitilde\textasciitilde[System ID of thread]\textasciitilde\textasciitilde} \texttt{ [Call Counter] [Image Name]![Native API Name]}  
In the context of Native APIs, the name of the image is always \texttt{ntdll}, but this is not the case when tracing APIs.

@@

\begin{itemize}
\item Firstly, it is determined if the argument being logged is a named constant, i.e., it is checked if \texttt{enum\_name} is different from \texttt{NULL}. In that case, the \texttt{get\_arg\_syname} function is invoked, which is in charge of, given the input named constant value, finding the corresponding name and recording it to the log. The named constant resolution is carried out by initially fetching the \texttt{struct} containing all the members of the enumeration the named constant belongs to, through the use of the aforementioned hash map. Then, it is determined if there is a perfect match between the argument value and one of the values of the \texttt{struct}'s entries. If this occurs, then the associated symbolic name is simply retrieved from the \texttt{struct}'s entry. Otherwise, a linear search is performed to unveil possible composite named constants (e.g., \texttt{FILE\_SHARED\_READ | FILE\_SHARED\_WRITE}).

@@

\begin{itemize}
\item Then, if the argument's type is primitive (e.g., \texttt{int}, \texttt{bool}, etc.),  \texttt{print\_simple\_value} is invoked to record its value. In particular, the second parameter of this function is a boolean which is set to \texttt{true} if it is wanted to print the argument's value with leading zeros (e.g., if the argument is of pointer type and the address it stores is being logged ) and \texttt{false} otherwise. The way \texttt{print\_simple\_value} operates is quite straightforward:
\begin{enumerate}
\item At the beginning the argument's value is simply logged, with leading zeroes or not depending on the previously mentioned parameter
\item Then, if the argument is a pointer, it is determined if the value it points to needs to be logged. This occurs if the argument is being processed before the Native API execution (i.e., \texttt{pre} is \texttt{true}) and is an input parameter (i.e., \texttt{SYSARG\_READ} is contained in \texttt{flags}) or if the argument is being processed after the Native API execution (i.e., \texttt{pre} is \texttt{false}) and is an output parameter (i.e., \texttt{SYSARG\_WRITE} is contained in \texttt{flags}). The pointed value is obtained through the use of \texttt{PIN\_SafeCopy}, which copies the specified number of bytes from a source memory region to a destination memory region, guaranteeing safe return to the caller even if such regions are inaccessible.  
\end{enumerate}

@@

Once the input argument values are logged, then the reference to the \texttt{syscall\_tracer struct} representing the Native API under analysis is stored in thread local data (\textit{Listing 3.1}). This is done so that, after the Native API's execution, the corresponding \texttt{syscall\_tracer} data structure can be accessed from \texttt{TraceSysCallExit}. In fact, as mentioned in Section~\ref{ThreadMan}, the \texttt{bluepill\_tls struct} associated to the running thread can be retrieved in any analysis function by means of \texttt{PIN\_GetThreadData}.

@@

\section{Tracing Windows Libraries}

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamically linked libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g., file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\label{API Tracing}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure~\ref{API Tracing}}.

@@

\subsection{Instrumentation} \label{Instrumentation}

The first step taken towards APIs tracing is APIs instrumentation, i.e., placing analysis routines before and after the APIs' execution. To do this, a specific function was embedded in BluePill's image callback (the notion of image callback was previously introduced in Section~\ref{MainImage}). Such function, named \texttt{AddAnalysisRoutines}, performs the following actions for every DLL being loaded:

\begin{enumerate} 
\item It inspects the symbol table, i.e., the DLL's exports. To this end, Pin symbol objects (\texttt{SYM}s) are employed, which provide information about function symbols in applications \cite{Pin}.
\item For every symbol object, it retrieves the corresponding routine (\texttt{RTN}) object. In fact, in Pin a \texttt{RTN} represents a function usually generated by a compiler for a procedural programming language, e.g C \cite{Pin}. To retrieve the aforementioned \texttt{RTN} object, the \texttt{RTN\_FindByAddress} API was utilized, which takes as input a \texttt{RTN} memory address and returns a handle to the found \texttt{RTN}. Such memory address was determined by adding the symbol's RVA (obtained via \texttt{SYM\_Value}) to the lowest image address (fetched through \texttt{IMG\_LowAddress}).  
\item For every routine object, it retrieves the corresponding name via \texttt{RTN\_name} and checks whether the API being represented is "known", i.e., a source of information associated to the API exists.

@@

\textit{drltrace} (already mentioned in Section~\ref{DBI}) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry on \textit{Listing 3.9}.      
\vspace{1cm}
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

Before API instrumentation takes place, BlueTracer parses the aforementioned configuration file and, for every entry, builds a \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) for each argument in the entry. We decided to reuse \texttt{sysinfo\_arg\_t struct} since its fields were fitting with respect to the arguments information which had to be stored. Therefore, after this parsing process, each configuration file entry is represented by an <API name, \texttt{vector} of \texttt{sysinfo\_arg\_t}> pair. Such pairs are then utilized to build a hash map, which maps the API name to the corresponding arguments. This hash map is used in step 3 of the APIs instrumentation process to, given the API name, fetch the appropriate information. In particular, it is important to note that, from a performance point of view, it is much better to include the information retrieval operations in the image instrumentation function, rather than in the analysis ones, as the former are executed just once for every image, while the latter are invoked each time an API is called by the pinned application.

The other source of API information employed by BlueTracer is \textit{PyREbox}'s database. \textit{PyREbox} is a Python scriptable reverse engineering sandbox based on QEMU and VMI techniques, whose goal is to provide support for reverse engineering through its dynamic analysis capabilities \cite{Pyrebox}. \textit{PyREbox} is equipped with its own API tracer, which relies on the information contained in a sqlite database to correctly log the APIs' parameters. Such database was, in turn, generated by utilizing as its core the data from the \textit{Deviare}\footnote{Deviare is an open-source hooking engine for instrumenting arbitrary \texttt{Win32} functions \cite{Deviare}} project and then employing a crawler for parsing the Microsoft online documentation with the goal of marking which parameters are input parameters and which are output parameters. The resulting database is incredibly richer in API information than \textit{drltrace}'s configuration file. In fact, the number of APIs for which data is stored is much higher, argument names are also included, as well as the contents of named constants' enumerations, information which \textit{drltrace}'s configuration file does not provide at all.

@@

The most important field of such \texttt{struct} is surely the arguments' array, which is made up of \texttt{libcall\_arg\_info\_t} data structures (\textit{Listing 3.11}) containing the information relative each argument of an API, including an entry specific for the return value data. The array size was set to 19, since it was observed from the database data that the API with the most parameters had 18 arguments. This time, we decided not to reuse NTAPI's \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) to store the API's arguments information due to its lack of the field storing the argument name. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing information associated to an API argument},captionpos=b]
typedef struct _libcall_arg_info_t {
	int arg_num;				// Arg Ordinal
	const char* arg_name;		// Arg Name
	int arg_type;             	// Arg Type
	char* arg_type_name;		// Arg Type Name
	int size;					// Size
	bool in_out_flag;			// Input/Output Flag
} libcall_arg_info_t;
\end{lstlisting}

@@

Due to the high amount of APIs for which information was stored in the \textit{PyREbox} database, we decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second-level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure~\ref{Hashmap}}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\label{Hashmap}
\end{figure}

@@

\subsection{Analysis before the call}
